---
title: "Making Functions in R"
description: |
  Examples of functions made in R using the case_when() function an a while loop
author:
  - name: Jaleise Hall
    url: {}
date: 03-16-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Example Function Using case_when()

Functions in R a created using .R scripts. In order to comment out text in the .R file, you would use # just as you would in an R markdown document. 

```{r}
# CASE_WHEN: for multiple conditions

# Contract: Classifies plastic types as PETE, PE, PVC, PP, PS, and OTHER based on the chemical composition of a sample of the plastic in question.

detect_plastic <- function(C, H, O, Cl) {
  
  # error checking
  C = ifelse(
    C < 0,
    return("Carbon content must be a value of 0 or greater than 0. Check for NA values"),
    C
  )
  
  H = ifelse(
    H < 0,
    return("Hydrogen content must be a value of 0 or greater than 0. Check for NA values"),
    H
  )
  
  O = ifelse(
    O < 0,
    return("Oxygen content must be a value of 0 or greater than 0. Check for NA values"),
    O
  )
  
  Cl = ifelse(
    Cl < 0,
    return("Chlorine content must be a value of 0 or greater than 0. Check for NA values"),
    Cl
  )
  
  # case_when loop
  detect_plastic = case_when(
    C == 10 & H == 8 & O == 4 & Cl == 0 ~ "PETE",
    C == 8 & H == 8 & O == 0 & Cl == 0 ~ "PS",
    C == 3 & H == 6 & O == 0 & Cl == 0 ~ "PP",
    C == 2 & H == 4 & O == 0 & Cl == 0 ~ "PE",
    C == 2 & H == 3 & O == 0 & Cl == 1 ~ "PVC",
    TRUE ~ "other"
  )
}

```

Setting up functions can be made easier by inserting a roxygen skeleton through the RStudio path: Code > Insert Roxygen Skeleton. In order to inser the skeleton, you must have your cursor inside the function you are trying to create.

### Example Function Using a While Loop

```{r}
# WHILE: for repeating until a condition is met

# Contract: How many more wastewater treatment plants (globally (US) must adopt tertiary treatment as part of their treatment process for oceanic plastic toxicity by WWTP outfall into surrounding oceans to decrease by 20% if there are currently 6,000 US WWTPs utilizing tertiary treatment.

# # of WWTPs in US: 14,748 (2017)
# Current tertiary: 6,000 (made up value)
# current toxicity: 900 kg/AF
# % decrease: 20%
# toxicity goal: 720 kg/AF
# addition of 1 plant using tertiary causes: 0.03% decrease in toxicity

tertiary_plants <- function(tox = 1, tert = 0) {

  # error checking
  tox = ifelse(
    tox > 720,
    return("Plastic toxicity must be less than 720"),
    tox
  )
  
  tert = ifelse(
    tert > 8748,
    return("Number of plants must be less than 8748"),
    tert
  )
  
  # initial 
  toxicity = tox
  tertiary = tert
  
  # while loop
  while((tertiary < 8748) &&(toxicity < 720)) {
    toxicity = 900*(tertiary*0.0003)
    tertiary = tertiary + 1
    toxicity = toxicity + 0.27
    print(tertiary)
    print(toxicity)
  }
  # return the number of plants that must adopt tertiary treatment
  return(tertiary)
}


```


Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.


