---
title: "Cetacean species richness"
description: |
  Mapping species richness of cetacean species in the coastal California area, California Bight.
author:
  - name: Jaleise Hall
    url: {}
date: 02-26-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(raster)
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(fasterize)
library(rnaturalearth)
```

### Create Raster

```{r}

rastlist <- list.files(path = here::here("project_tab_data", "ca_cetaceans"), full.names=TRUE)

allrasters <- lapply(rastlist, raster)

ca_cetaceans <- 
  raster::stack(rastlist)

plot(ca_cetaceans)

threshold <- 0.8

# Reclassification matrix 
rec_cet_matrix <- c(-Inf, threshold, 0,
                   threshold, Inf, 1,
                   0, threshold, 0,
                   0, Inf, 0)
# Reclassify 
rec_cetaceans <- reclassify(ca_cetaceans, rcl = rec_cet_matrix)

# Convert to data frame
cetaceans_df <- raster::rasterToPoints(rec_cetaceans) %>%
  as.data.frame() %>% 
  clean_names() %>% 
  pivot_longer(cols = balaenoptera_acutorostrata:ziphius_cavirostris, names_to = "species", values_to = "presence_code")

cetaceans_df[is.na(cetaceans_df[])] <- 0

cetaceans_presence <- cetaceans_df %>% 
   mutate(presence_status = case_when(
    presence_code == 1 ~ "present",
    presence_code == 0 ~ "not present"
  ))

```

### Plot the Raster

```{r}

# Plot reclassified layer
# plot(depth_binary)

world <- ne_countries(scale = "medium", returnclass = "sf")
world_coastline <- ne_coastline(scale = "medium", returnclass = "sf")


cetacean_richness <- ggplot() +
  geom_raster(data = cetaceans_presence, aes(x = x, y = y, fill = presence_status)) +
  geom_sf(data = world_coastline) +
  geom_sf(data = world) +
  coord_sf(xlim = c(-125, -115), ylim = c(32, 38)) +
  scale_fill_manual(values = c("deeppink3", "mediumorchid1")) +
  labs(fill = "",
       x = "longitude",
       y = "latitude", 
       title = "California Bight Cetacean Species Richness HALL 2020") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(here("project_tab_data", "cetacean_richness.png"), height = 4, width = 6)

```


![**Figure 1.** This map shows the cetacean species richness for the 35 cetacean species found in the California Bight. This richness is reflective of probability of occurance based on species preferences for water temperature, depth, salinity, and distance to land. Presence was determined based on an occurance threshold of 0.8 such that if occurance was at or about 0.8, then the species would be deemed "present".](cetacean_richness.png)

Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.


